---
title: "NOAA Storm DB Analysis"
author: "nuno r"
date: "January 28, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

Document Layout
Language: Your document should be written in English.
Title: Your document should have a title that briefly summarizes your data analysis
Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis 
         in at most 10 complete sentences.

There should be a section titled Data Processing which describes (in words and code) how the data were loaded 
into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the 
data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider 
using the cache = TRUE option for certain code chunks.

There should be a section titled Results in which your results are presented.
You may have other sections in your analysis, but Data Processing and Results are required.
The analysis document must have at least one figure containing a plot.

Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), 
but there cannot be more than three figures total.
You must show all your code for the work in your analysis document. This may make the document a bit verbose, 
but that is okay. In general, you should ensure that echo = TRUE for every code chunk (this is the default 
setting in knitr).


Your data analysis must address the following questions:
  
Q1) Across the United States, which types of events (as indicated in the EVTYPE variable) 
   are most harmful with respect to population health?

Q2) Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be 
responsible for preparing for severe weather events and will need to prioritize resources for different types 
of events. However, there is no need to make any specific recommendations in your report.

## Introduction
This report analizes data from NOAA Storm DB and identifies the impacts on health and damages to properties and crops.
We will demonstrate the top 10 events with the most impact on health and an aggregate of all damages to properties and crops.
We expect anyone reading this document will be able to analyze the data and process it to reach similar results.

## Data Processing
Let's load and prepare the data for analysis.

Load dplyr to manipulate data
```{r}
# Additional information can be found here:
# https://dplyr.tidyverse.org/
library(plyr)
library(dplyr)
```

Load ggplot2 to create the charts
```{r}
# Additinal information can be found here:
# https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
library(ggplot2)
```

Load the NOAA DB file directly into the workspace with no external manipulations
```{r}
# NOTE: The original NOAA DB file can be found here:
# https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
noaa_db_url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(noaa_db_url, dest = "StormData.csv.bz2")
```

Read the Storm Data DB extract into a table
```{r}
# NOTE: additional info on read.table:
# https://stat.ethz.ch/R-manual/R-devel/library/utils/html/read.table.html 
noaa_table <- read.table("StormData.csv.bz2", 
                         sep = ",", 
                         header = TRUE, 
                         quote = "\"'", 
                         dec = ".",
                         numerals = c("allow.loss", 
                                      "warn.loss", 
                                      "no.loss"))

# Find the names of the columns in the dataset
names(noaa_table)
```

We don't need all columns. Pick the necessary columns for further analysis
```{r}
accidents <- noaa_table[,c(8,23:24)] # include EVTYPE, FATALITIES, INJURIES
property_damages <- noaa_table[,c(8,25:28)] # include EVTYPE, PROPFMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
```

Used only for debugging, too much information otherwise
```{r}
#NOTE: used only for debugging, too much information otherwise
#print(accidents)
#print(property_damages)
```

Identify the highest incidence of events, sorted by fatalities and injuries that did not result in a fatality
```{r}
top_display_events = 10
top.accidents <- aggregate(cbind(INJURIES,FATALITIES) ~ EVTYPE, 
                           data = accidents, 
                           sum, 
                           na.rm=TRUE)

top.accidents <- arrange(top.accidents, 
                         desc(INJURIES + FATALITIES))

# pick the top event types
top.accidents <- top.accidents[1:top_display_events,]
top.accidents
```

## Results
### Q1) Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Prepare the fatalities data to be visualized
```{r}
fatalities_chart = ddply(top.accidents, 
                         .(EVTYPE), 
                         summarize, 
                         sum_fatalities = sum(FATALITIES,na.rm=TRUE))

fatalities_chart = fatalities_chart[order(fatalities_chart$sum_fatalities, decreasing = TRUE), ]
head(fatalities_chart,top_display_events)
```

Events that resulted in fatalities
```{r}
ggplot(fatalities_chart[1:top_display_events, ], 
       aes(EVTYPE,
           x = reorder(top.accidents$EVTYPE, sum_fatalities),
           y = sum_fatalities,
           fill=EVTYPE,
           alpha=0.5
          )
) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = fatalities_chart$sum_fatalities), 
            size = 5, 
            hjust = 0.5, 
            vjust = 1, 
            position = "stack") +
  ggtitle("Events = Fatalities") +
  guides(color = "none") +
  coord_flip() +
  xlab("") + 
  ylab("") +
  scale_fill_grey() + 
  theme_classic()

ggsave("fatalities-1.png")
dev.off()
```

Prepare the injuries data to be visualized
```{r}
injuries_chart = ddply(top.accidents, 
                       .(EVTYPE), 
                       summarize, 
                       sum_injuries = sum(INJURIES,na.rm=TRUE))
injuries_chart = injuries_chart[order(injuries_chart$sum_injuries, decreasing = TRUE), ]
head(injuries_chart,top_display_events)
```

Events that resulted in injuries
```{r}
ggplot(injuries_chart[1:top_display_events, ], 
       aes(EVTYPE,
           x = reorder(top.accidents$EVTYPE, sum_injuries),
           y = sum_injuries,
           fill=EVTYPE,
           alpha=0.5,
           label = sum_injuries
       )
) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = fatalities_chart$sum_fatalities), 
            size = 5, 
            hjust = 0.5, 
            vjust = 1, 
            position = "stack") +
  ggtitle("Events = Injuries") +
  guides(color = "none") +
  coord_flip()+
  xlab("") + 
  ylab("") +
  scale_fill_grey() + 
  theme_classic()

ggsave("injuries-1.png")
dev.off()
```

### Q2) Across the United States, which types of events have the greatest economic consequences? Find out property damage and crop damages

Prepare the damages data to be visualized. This data includes property damages and crop damages
```{r}
top.damages <- aggregate(cbind(PROPDMG,CROPDMG) ~ EVTYPE, 
                           data = property_damages, 
                           sum, 
                           na.rm=TRUE)

top.damages <- arrange(top.damages, 
                       desc(PROPDMG + CROPDMG))

top.damages <- top.damages[1:top_display_events,]
top.damages
```

Find out how to apply the multiplier for PROPDMGEXP and CROPDMGEXP
```{r}
table(property_damages$PROPDMGEXP)
table(property_damages$CROPDMGEXP)
```

Property multipliers
```{r}
property_damages$PROPDMGCALC [property_damages$PROPDMG==0] <- 0  
property_damages$CROPDMGCALC [property_damages$CROPDMG==0] <- 0 

property_damages$PROPDMGCALC [property_damages$PROPDMGEXP=="H" | 
                                property_damages$PROPDMGEXP=="h"] <- property_damages$PROPDMG[property_damages$PROPDMGEXP=="H" |
                                                                                                property_damages$PROPDMGEXP=="h"]*100
property_damages$PROPDMGCALC [property_damages$PROPDMGEXP=="K" | 
                                property_damages$PROPDMGEXP=="k"] <- property_damages$PROPDMG[property_damages$PROPDMGEXP=="K" | 
                                                                                                property_damages$PROPDMGEXP=="k"]*1000

property_damages$PROPDMGCALC [property_damages$PROPDMGEXP=="M" | 
                                property_damages$PROPDMGEXP=="m"] <- property_damages$PROPDMG[property_damages$PROPDMGEXP=="M" |
                                                                                                property_damages$PROPDMGEXP=="m"]*1000000

property_damages$PROPDMGCALC [property_damages$PROPDMGEXP=="B" | 
                                property_damages$PROPDMGEXP=="b"] <- property_damages$PROPDMG[property_damages$PROPDMGEXP=="B" | 
                                                                                                property_damages$PROPDMGEXP=="b"]*1000000000
```

CROP multipliers
```{r}
property_damages$CROPDMGCALC [property_damages$CROPDMGEXP=="H" | 
                                property_damages$CROPDMGEXP=="h"] <- property_damages$CROPDMG[property_damages$CROPDMGEXP=="H" | 
                                                                                                property_damages$CROPDMGEXP=="h"]*100
property_damages$CROPDMGCALC [property_damages$CROPDMGEXP=="K" | 
                                property_damages$CROPDMGEXP=="k"] <- property_damages$CROPDMG[property_damages$CROPDMGEXP=="K" | 
                                                                                                property_damages$CROPDMGEXP=="k"]*1000

property_damages$CROPDMGCALC [property_damages$CROPDMGEXP=="M" | 
                                property_damages$CROPDMGEXP=="m"] <- property_damages$CROPDMG[property_damages$CROPDMGEXP=="M" | 
                                                                                                property_damages$CROPDMGEXP=="m"]*1000000
property_damages$CROPDMGCALC [property_damages$CROPDMGEXP=="B"| 
                                property_damages$CROPDMGEXP=="b"] <- property_damages$CROPDMG[property_damages$CROPDMGEXP=="B" |
                                                                                                property_damages$CROPDMGEXP=="b"]*1000000000
```

Find total damage
```{r}
total_damages <- aggregate(cbind(PROPDMGCALC,CROPDMGCALC) ~ EVTYPE, 
                         data = property_damages, 
                         sum, 
                         na.rm=TRUE)

total_damages <- arrange(total_damages, 
                       desc(PROPDMGCALC + CROPDMGCALC))

total_damages <- total_damages[1:top_display_events,]
total_damages

total_damages_sum = total_damages$PROPDMGCALC + total_damages$CROPDMGCALC
total_damages_sum

total_damages_table = as.data.frame(cbind(total_damages, total_damages_sum))
total_damages_table
```

Total $ amount of damages to properties and crops resulting from weather-related events
```{r}
ggplot(total_damages[1:top_display_events, ], 
       aes(EVTYPE,
           x = reorder(total_damages$EVTYPE, total_damages_sum),
           y = total_damages_sum,
           fill=EVTYPE,
           alpha=0.5,
           label = total_damages_sum
       )
) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = total_damages_sum), 
            size = 5, 
            hjust = 0.5, 
            vjust = 1, 
            position = "stack") +
  ggtitle("Total Damages = Property + Crops ($)") +
  guides(color = "none") +
  coord_flip()+
  xlab("") + 
  ylab("") +
  scale_fill_grey() + 
  theme_classic()

ggsave("damages_costs-1.png")
dev.off()
```

## Conclusion and analysis of results
After exploring the data extensively we were able to identify FLOODS as the most impactful from a an economy point of view. TORNADOS are the most devastating weather-related event, capable of causing highest number of injuries and fatalities.